Reproducible development environments with rix

### 2025-02-20 paper.qmd
done?

While pinning revisions is essential for reproducibility, this is not the only
reason [Nix]{.pkg} is very useful for reproducibility, but the [Nix]{.pkg}
package manager itself was designed with reproducibility in mind. [Nix]{.pkg} is
a functional package manager, meaning that it uses concepts from functional
programming to build software: no-side effects or global variables are allowed,
and the output is always the same if given the same inputs, regardless of where
or when a package gets built (an important side-note is in order here: while the
functional aspect of [Nix]{.pkg} is quite useful to ensure reproducibility, it
can lead to situations where writing the required [Nix]{.pkg} expression to
build it can be much more difficult than usual: this is especially true for
packages that must download assets from the internet during installation, such
as several BioConductor packages like [musData]{.pkg} for example. But this is
only a concern for the packagers, not users of the package that simply wish to
install and use the package). [Nix]{.pkg} also allows for multiple versions (or
variants in the [Nix]{.pkg} jargon) to be installed at the same time on the same
system. Each variant of a package has a unique identifier, so that it is
possible to have two versions of [R]{.proglang} installed and use the correct
one where adequate. We refer to @dolstra2004nix for a more technical discussion
on [Nix]{.pkg}.

@@

The first thing this fork allows us to do is offer a daily snapshot of CRAN.
Every day, thanks to Github Actions, the [R]{.proglang} packages set gets
updated, and the result commited to a dated branch. This has been going on since
the 14th of December 2024 (see
[here](https://github.com/rstats-on-nix/nixpkgs/tree/2024-12-14)). So when users
set a date as in `rix(date = "2024-12-14", ...)` the specified dated branch from
the fork is going to get used.

But this doesn’t mean that we recommend you
use any date from the `rstats-on-nix/nixpkgs` fork: instead, each Monday,
another action uses this fork and tries to build a set of popular packages on
Linux and macOS, and only if this succeeds is the date added through a PR to the
list of available dates on `{rix}`!

The reason this is done like this is to manage another *risk* of the upstream
`nixpkgs`. As you know, `nixpkgs` is huge, and though the utmost care is taken
by contributors and the PR review process is very strict, it can happen that
updating packages breaks other packages. For example recently RStudio was in a
broken state due to an issue in one its dependencies, `boost`. This is not the
fault of anyone in particular: it's just that packages get updated and packages
that depend on them should get updated as well: but if that doesn't happen
quickly enough, the `nixpkgs` maintainer faces a conundrum. Either he or she
doesn't update the package because it breaks others, but not updating a package
could be a security vulnerability, or he or she updates the package, but now
others, perhaps less critical packages are broken and need to be fixed, either
by their upstream developers, or by the `nixpkgs` maintainer of said packages.
In the case of RStudio a fix was proposed and promptly merged, but if you wanted
to install RStudio during the time it took to fix it, you would have faced
an error message, which isn't great if all you want is use [Nix]{.pkg} shells as
development environments.

So for us, having a fork allows us to backport these fixes and so if you try to
install RStudio using the latest available date, which is `"2025-02-10"`, it’s
going to work, whereas if you tried to build it on that date using upstream
`nixpkgs` you'd be facing an error!

We spent quite some time backporting fixes: we went back all the way to
2019. The way this works, is that we start by checking out a `nixpkgs` commit on
selected dates, then we "update" the [R]{.proglang}packages set by using the Posit CRAN and
Bioconductor daily snapshots. Then, we backport as many fixes as possible, and
ensure that a selection of popular packages work on both x86-linux (which
includes Windows, through WSL) and aarch64-darwin (the M-series of Macs). Then
we commit everything to a dated branch of the `rstats-on-nix/nixpkgs` fork. You
can check out all the available dates by running: `rix::available_dates()`.
We're pretty confindent that you should not face any issues when using [Nix]{.pkg} to
build reproducible environments for R. However, should you face a problem, don't
hesitate to open an issue!

We have now packages and [R]{.proglang}versions working on Linux and macOS from March 2019
to now. See [this
repository](https://github.com/rstats-on-nix/daily_cran/blob/master/readme.md)
that contains the scripts that allowed us to do it. Backporting fixes was
especially important for Apple Silicon computers, as it took some time for this
platform to work correctly on Nix. By backporting fixes, we can now provide
olders versions of these packages for Apple Silicon as well!

Using this approach, our fork now contains many more versions of working R
packages than upstream. [rix]{.pkg} will thus likely keep pointing towards our fork
in the future, and not upstream anymore. This should provide a much better user
experience. An issue with our fork though, is that by backporting fixes, we
essentially create new [Nix]{.pkg} packages that are not included in upstream, and thus,
these are not built by Hydra, Nix's CI platform which builds binary packages. In
practice this means that anyone using our fork will have to compile many
packages from source. Now this is pretty bad, as building packages from source
takes quite some time. But fear not, because thanks to
[Cachix](https://www.cachix.org/) we now also have a dedicated binary cache of
packages that complements the default, public [Nix]{.pkg} cache! We provide instructions
on how to use Cachix, it's very easy, it's just running 2 additional commands
after installing Nix. Using Cachix speeds up the installation process of
packages tremendously. I want to give my heartfelt thanks to
[Domen Kožar](https://www.cachix.org/about) for sponsoring the cache!

Another thing we do with our fork is run an action every day at midnight, that
monitors the *health* of the [R]{.proglang}packages set. Of course, we don't build every
CRAN package, merely a handful, but these are among the most popular or the most
*at-risk* of being in a broken state. See
[here](https://github.com/rstats-on-nix/monitor_health/actions).

@@

If one wishes to run pipelines written using the [targets]{.pkg} package
[@landau2021] on Github Actions, [rix]{.pkg} features a function called
`tar_nix_ga()`, inspired by the `tar_github_actions()` function from the
[targets]{.pkg} package. This function writes a YAML file under
`.github/workflows/` which then runs the [targets]{.pkg} pipeline using Github
Actions at each push, in a completely reproducible environment. For such an
example, consult the code in this
repository^[https://github.com/b-rodrigues/rix_cicd].

The final feature that we provide is the ability to run one arbitrary function
from what we call a *subshell*. Suppose that you are running the latest version
of R, and that you wish or need to run one particular function from an older
release of R, or from an older package. Using the `with_nix()` function, it is
possible to run this single function inside the right environment, and recover
its output in the main session. To illustrate, consider the following two lines
of code:

@@

On versions of [R]{.proglang}released after version 3.6, these two lines return the
following vector: `r out_r`, but in versions prior to 3.6, the result will be
different. This is a documented change as shown in the [R]{.proglang}version 3.6 release
notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]

@@

It is possible to use `with_nix()` from an [R]{.proglang}session that was installed itself
via Nix, or from a regular [R]{.proglang}installation.

@@

There are many tools to make one’s analyses reproducible, but [Nix]{.pkg} distinguishes
itself from these in that its main concern is in deploying the complete set of
packages that are *closed under the ''depends on'' relation*: [Nix]{.pkg} does not
simply install one piece of software, but its dependencies, and their
dependencies. This makes [Nix]{.pkg} extremely useful for reproducible research, but
because it solves a complex problem, it is a complex tool. With [rix]{.pkg}, we
wish to make using [Nix]{.pkg} more accessible by providing a familiar interface and
user experience to [R]{.proglang} users. Because it is possible to build
reproducible development shells using  [Nix]{.pkg}, it can accomodate any
use case: running scripts, pipelines, interactive [shiny]{.pkg} applications,
and serve [plumber]{.pkg} application programming interfaces.

>>>>

While pinning revisions is essential for reproducibility, this is not the only
reason [Nix]{.pkg} is very useful for reproducibility, but the [Nix]{.pkg}
package manager itself was designed with reproducibility in mind. [Nix]{.pkg} is
a functional package manager, meaning that it uses concepts from functional
programming to build software: no-side effects or global variables are allowed,
and the output is always the same if given the same inputs, regardless of where
or when a package gets built (an important side-note is in order here: while the
functional aspect of [Nix]{.pkg} is quite useful to ensure reproducibility, it
can lead to situations where writing the required [Nix]{.pkg} expression to
build it can be much more difficult than usual: this is especially true for
packages that must download assets from the internet during installation, such
as several Bioconductor packages like [musData]{.pkg} for example. But this is
only a concern for the packagers, not users of the package that simply wish to
install and use the package). [Nix]{.pkg} also allows for multiple versions (or
variants in the [Nix]{.pkg} jargon) to be installed at the same time on the same
system. Each variant of a package has a unique identifier, so that it is
possible to have two versions of [R]{.proglang} installed and use the correct
one where adequate. We refer to @dolstra2004nix for a more technical discussion
on [Nix]{.pkg}.

@@

The first thing this fork allows us to do is offer a daily snapshot of CRAN.
Every day, the [R]{.proglang} packages set gets updated, and the result commited
to a dated branch using scripts running on GitHub Actions (a continuous
integration and deployment platform). This has been going on since the 14th of
December 2024 (see
[here](https://github.com/rstats-on-nix/nixpkgs/tree/2024-12-14)). So when users
set a date as in `rix(date = "2024-12-14", ...)` the specified dated branch from
the fork is going to get used.

But this doesn’t mean that we recommend users use any date from the
`rstats-on-nix/nixpkgs` fork: instead, each Monday, another action uses this
fork and tries to build a set of popular packages on Linux and macOS, and only
if this succeeds is the date added through a PR to the list of available dates
on `{rix}`. This means that we also monitor the health of popular [R]{.proglang}
packages and can propose a fix upstream if a package stops building
successfully, which is something that can sometimes happen. For example,
recently RStudio was in a broken state due to an issue in one of its
dependencies, `boost`. This is not the fault of anyone in particular: it's just
that packages get updated and packages that depend on them should get updated as
well: but if that doesn't happen quickly enough, the `nixpkgs` maintainer faces
a conundrum. Either he or she doesn't update the package because it breaks
others, but not updating a package could be a security vulnerability, or he or
she updates the package, but now other, perhaps less critical packages are
broken and need to be fixed, either by their upstream developers, or by the
`nixpkgs` maintainer of said packages. In the case of RStudio a fix was proposed
and promptly merged, but if users wanted to install RStudio during the time it
took to fix it, they would have faced an error message.

However, thanks to our fork, we can back-port fixes and so if users try to
install RStudio using the date `"2025-02-10"` (which is when it was broken
on upstream `nixpkgs`), it'll work.

We spent quite some time backporting fixes: we went back all the way back to
March 2019. To do so, we start by checking out a `nixpkgs` commit on selected
dates, then we "update" the [R]{.proglang} packages set by using the Posit CRAN
and Bioconductor daily snapshots at that date. Then, we backport as many fixes
as possible, and ensure that a selection of popular packages work on both
x86-linux (which includes Windows, through WSL2) and aarch64-darwin (which is
how Apple Silicon Apple computers are designated as). Then we commit everything
to a dated branch of the `rstats-on-nix/nixpkgs` fork. Users can consult all the
available dates by running: `rix::available_dates()`.

An issue with forking `nixpkgs` though, is that by backporting fixes, we
essentially create new [Nix]{.pkg} packages that are not included in upstream,
and thus, these are not built by Hydra, Nix's CI platform which builds binary
packages. In practice this means that anyone using our fork will have to compile
many packages from source, which can be very time consuming. However, we also
provide a binary cache of packages sponsored by
[Cachix](https://www.cachix.org/) which complements the default, public
[Nix]{.pkg} cache. We provide instructions on how to use Cachix in the package's
Readme file. Using Cachix speeds up the installation process of packages
tremendously, as they simply get downloaded instead of being built (we build
a selection of packages only however, not the whole of CRAN and Bioconductor).

@@

If one wishes to run pipelines written using the [targets]{.pkg} package
[@landau2021] on Github Actions, [rix]{.pkg} features a function called
`tar_nix_ga()`, inspired by the `tar_github_actions()` function from the
[targets]{.pkg} package. This function writes a YAML file under
`.github/workflows/` which then runs the [targets]{.pkg} pipeline using Github
Actions on each push, in a completely reproducible environment. For such an
example, consult the code in this
repository^[https://github.com/b-rodrigues/rix_cicd].

The final feature that we provide is the ability to run one arbitrary function
in what we call a *subshell*. Suppose that you are running the latest version
of R, and that you wish or need to run one particular function from an older
release of R, or from an older package. Using the `with_nix()` function, it is
possible to run this single function inside the right environment, and recover
its output in the main session. To illustrate, consider the following two lines
of code:

@@

On versions of [R]{.proglang} released after version 3.6, these two lines return
the following vector: `r out_r`, but in versions prior to 3.6, the result will
be different. This is a documented change as shown in the [R]{.proglang} version
3.6 release notes.^[https://stat.ethz.ch/pipermail/r-announce/2019/000641.html]

@@

It is possible to use `with_nix()` from an [R]{.proglang} session that was
installed itself via Nix, or from a regular [R]{.proglang} installation.

@@

There are many tools to make one’s analyses reproducible, but [Nix]{.pkg}
distinguishes itself from these in that its main concern is in deploying the
complete set of packages that are *closed under the ''depends on'' relation*:
[Nix]{.pkg} does not simply install one piece of software, but its dependencies,
and their dependencies. This makes [Nix]{.pkg} extremely useful for reproducible
research, but because it solves a complex problem, it is a complex tool. With
[rix]{.pkg}, we wish to make using [Nix]{.pkg} more accessible to users of
[R]{.proglang} by providing a familiar interface and user experience. Because it
is possible to build reproducible development shells using [Nix]{.pkg}, it can
accomodate any use case: running scripts, pipelines, interactive [shiny]{.pkg}
applications, and serve [plumber]{.pkg} application programming interfaces.
